Searching for standards of fairness in the transportation justice literature

### 2024-04-14 Manuscript/Manuscript.qmd
Draft of journal-manuscript

---
title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
subtitle: "Searching for standards of fairness"
author:
  - name:
      given: Anastasia
      family: Soukhov
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: soukhoa@mcmaster.ca
  - name:
      given: Ignacio
      family: Tiznado-Aitken
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: i.tiznadoaitken@utoronto.ca
  - name:
      given: Matthew
      family: Palm
    affiliations:
      - name: University of North Carolina Chapel Hill
        department: Department of City and Regional Planning
    email: palmmatt@unc.edu
  - name:
      given: Steven
      family: Farber
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: steven.farber@utoronto.ca
  - name:
      given: Antonio
      family: Paez
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: paezha@mcmaster.ca
bibliography: All_papers_in_extracted.bib
format: docx
---

@@

<!--We suggest: Who and What: the tools of mobility. I like our more-->
**What: the tools of mobility.** Most studies deploying equity standards apply them to either transit (37%) or pedestrian modes (25%). Many transit analyses are multimodal in nature, comparing the access transit provides against the access automobiles provided as an implicit standard for comparison. Others combine transit with non-auto modes to examine access for households without vehicles. These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals. However, they often fail to include explicit minimums of service that should be provided to different populations. In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.

<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e. accessibility to specific destinations, or generally), with most studies focused on the latter. Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations. Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace. Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.

<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards. The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction. Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks. Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design. But infrastructure standards around sustainable infrastructure (i.e. block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses. The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:

@@

Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions. Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative. The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts. Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.

<!-- We suggested:
Call 3: Data availability as a matter of justice-->
Call 3: Data availability as a matter of justice.

Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations. Furthermore, data issues prevent the application of many methods of equity analyses in the Global South. The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act. In the Canadian context, Mobilizing Justice’s National Survey working group will advance data practices by conducting the first national survey on transportation poverty.

@@

Approximately 10% of the reviewed studies assessed interventions. Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice. Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations. A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve. Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.

@@

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity. In response to this focus, a lively and rapidly growing literature has emerged on the topic (see _inter alia_ [@martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]). A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered when the increase in Santiago Metro's subway fare sparked a period of massive demonstrations and unrest in October 2019 motivated by country-wide social inequalities; [@BBCNews2019; @diazpabonInequality2021]). More optimistically, in democratic systems, tracking objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. Several challenges arise when approaching this endeavor. The notorious complexity of transportation systems is one challenge: transport systems simultaneously move people, goods, and information, and emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020]. The benefits and burdens of transportation systems are diffuse over space and time. For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].

@@

While commendable, the strategies are too vague, which makes it possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity? What should that level be? What are the criteria for deciding that a community is "potentially affected"? What benefits should be distributed? Should the benefits be based on simple population weights? Or, contrariwise, should more deprived individuals be eligible for a larger share of the benefits?

These questions boil down to the development and use of _standards_ for transportation justice. The term "standard" connotes "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard]. How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom. For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower emission standards [@gouldsonFirms2006]. Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

> "_Man's capacity for justice makes democracy possible, but man's inclination to injustice makes democracy necessary._"
> ― Reinhold Niebuhr

> "_Never forget that justice is what love looks like in public_."
> ― Cornel West

@@

It is said that justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440].  Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations and over time. That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:

@@

**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006].

@@

It might be argued that all of the above touch on forms of distributive justice. Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and the distribution of tangible resources as burdens (e.g., fines as a punishment). Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

@@

The last concept that we discuss in this section is that of a standard. Briefly, standards are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size. In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

@@

In the following section we will discuss the analytical apparatus that we will use to interrogate the literature on equity standards in transportation.

@@

For this report, we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature. This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs of transportation "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity).]. According to -@jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?". Conventionally, discussions about justice have been aspatial, or rather, seen from the point of view of social space instead of geographical space, despite an early interest of geographers on the matter [@pirieSpatial1983]. The texture of the questions becomes more immediate and crisp when talking about transportation, which is inherently about space and time.

- **"Where?"**: Such questions traditionally relate to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group e.g., members of the same nation state [@jaggarPhilosophicalChallengesGlobal2009, p. 3]<!--, a notion that can be expanded to include all members of the human species (e.g., with the Universal Declaration of Human Rights) and nature (e.g., such as the Ecuadorian constitution, the first ever country to grant legal rights to nature [@akchurinConstructingRightsNature2015])-->. In the case of justice in transportation, the question of "where?" is paramount; as it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the definition of the spatial boundaries in the analysis is the answer to "where?".

@@

- When answering **"Who?"**, we inquire about which entities should be regarded as subjects or arbiters of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE - LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES PEOPLE WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of White men. A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward. Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control (e.g., captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]). In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of inequities, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of this potentials happens through travel. However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations. In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is fair, that is, (in)equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.

>>>>

---
title: "Searching for standards of fairness in the transportation justice literature"
format:
  pdf: default
  tandf-pdf:
    keep-tex: true

author:
  - name:
      given: Anastasia
      family: Soukhov
    affiliations: 
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: soukhoa@mcmaster.ca
  - name:
      given: Ignacio
      family: Tiznado-Aitken
    affiliations: 
      - name: University of Toronto
        department: Department of Geography and Planning
    email: i.tiznadoaitken@utoronto.ca  
  - name:
      given: Matthew
      family: Palm
    affiliations: 
      - name: The University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
    email: palmmatt@unc.edu  
  - name:
      given: Steven
      family: Farber
    affiliations: 
      - name: University of Toronto
        department: Department of Geography and Planning
    email: steven.farber@utoronto.ca  
  - name:
      given: Antonio
      family: Páez
    affiliations: 
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: paezha@mcmaster.ca
bibliography: bibliography.bib
---

@@

<!--We suggest: Who and What: the tools of mobility. I like our more-->
**What: the tools of mobility.** Most studies that deploy equity standards apply them to either transit (37%) or pedestrian modes (25%). Many transit analyses are multimodal in nature, comparing accessibility by transit to accessibility by automobiles as an implicit standard. Others combine transit with non-auto modes to examine access for households without vehicles. These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals. However, they often fail to include explicit minima of service that should be provided to different populations. In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.

<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e. accessibility to specific destinations, or generally), with most studies focused on the latter. Over a quarter of the papers that focus on accessibility do not consider specific destinations, but instead concentrate on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations. Other studies are centred on particular destinations considered as essential, including employment, healthcare, grocery stores, and greenspace. Access to community support or places of worship, as well as childcare activities are far less common in the literature reviewed.

<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
**How: concepts and standards of justice.** The most common standards found in the literature, which we term opportunity standards, are defined using cut offs in access to opportunities (37.2%). Almost as common is the formulation of standards that use thresholds related to the perceptions or activities of specific populations (35.8%); we call these population standards. The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction. Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population ones are evaluated through well-being frameworks. Another category of standards that emerged from the literature are based on infrastructure, and these include adherence to universal design guidelines or specific benchmarks for road network characteristics. Standards of this kind are comparatively less common in the literature and are often framed in terms of rights, particularly as they relate to achieving universal design. Infrastructure standards around sustainable infrastructure (i.e. block lengths or distance between transit stops) are also evaluated through vertical, horizontal and spatial lenses. The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:

@@

Mixed-methods research can help governments set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, as well as freedom of movement and perceived access, among other facets of transportation. Yet most of the research reviewed is purely quantitative or, to a much lesser degree, purely qualitative. The frequent deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts. Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.

<!-- We suggested:
Call 3: Data availability as a matter of justice-->
Call 3: Making data available is a matter of justice.

Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations. Furthermore, data issues prevent the application of many methods of equity analyses in the Global South. The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act. In the Canadian context, Mobilizing Justice’s National Survey working group is working to advance data practices by conducting the first national survey on transportation poverty and striving to make data available in the form of Open Data Products.

@@

Approximately 10% of the studies reviewed aimed to assess interventions. Delivering new equity-focused policies and programs that do not actually advance equity harms communities, generates mistrust, and delegitimates future efforts to promote transportation justice. Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations. A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve. Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.

@@

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity. In response to this focus, a lively and rapidly growing literature has emerged on the topic (see _inter alia_ [@martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]). A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered to their woe when an increase in Santiago Metro's fares became a flashpoint for social inequalities, and sparked a period of massive demonstrations and unrest in October 2019; [@BBCNews2019; @diazpabonInequality2021]). More optimistically, in democratic systems, tracking objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. Several challenges arise when approaching this endeavor. The notorious complexity of transportation systems is one of them: transport systems simultaneously move people, goods, and information. Emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020], turning both users and service providers into digital rentiers [@birchBig2022]. The benefits and burdens of transportation systems are diffuse over space and time. For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].

@@

While commendable, the strategies are too vague, which means it is possible to implement them in a myriad ways, either genuinely to comply with the spirit of justice, or else performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity greater than zero? What should that level be? What are the criteria for deciding that a community is "potentially affected"? What benefits should be distributed? Should the benefits be based on simple population weights? Or, contrariwise, should more deprived individuals be eligible for a larger share of the benefits?

These questions boil down to the development and use of _standards_ for transportation justice. The term "standard" connotes "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard]. How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom. For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower emission standards [@gouldsonFirms2006]. Regardless of the cause, the result is the same: pollution tends to be worse where poorer people are [@delucaCode2012].

Supporting the creation of (more) just transportation systems involves understanding the production and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief to planners, especially in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

@@

It is said that justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440].  Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations and over time. That said, it is possible to distinguish several forms of justice, including those listed below [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020].

@@

**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers, and the community according to moral values [@tylerRestorativeJusticeProcedural2006].

@@

It might be argued that all of the above touch on forms of distributive justice. Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and/or the distribution of tangible resources as burdens (e.g., fines as a punishment). Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

@@

The last concept that we discuss in this section is that of a standard. Briefly, standards are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group gives or receives in proportion to their size. In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

@@

In the following section we discuss the analytical apparatus that we use to interrogate the literature on equity standards in transportation.

@@

For this report, we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature. This is done either explicitly, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity; or implicitly, as in @gosslingUrban2016, who asks of the outputs of transportation "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity).]. According to @jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?". Conventionally, discussions about justice have been aspatial, or rather, seen from the point of view of social space instead of geographical space, despite an early interest of geographers on the matter [@pirieSpatial1983]. The texture of the questions becomes more immediate and crisp when talking about transportation, which is inherently about space and time.

- **"Where?"**: Such questions traditionally relate to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group e.g., members of the same nation state [@jaggarPhilosophicalChallengesGlobal2009, p. 3]<!--, a notion that can be expanded to include all members of the human species (e.g., with the Universal Declaration of Human Rights) and nature (e.g., such as the Ecuadorian constitution, the first ever country to grant legal rights to nature [@akchurinConstructingRightsNature2015])-->. In the case of justice in transportation, the question of "where?" is paramount, as it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the definition of the spatial boundaries in the analysis is the answer to "where?".

@@

- When answering **"Who?"**, we inquire about which entities should be regarded as subjects or arbiters of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE - LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES PEOPLE WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of White men. A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward. Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the mobility tools that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control [e.g., captive users of a single mode @jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to elucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of inequities, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of these potentials happen through travel. However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach places where activities that the traveler values take place. In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is fair, that is, (in)equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, or Sufficientarianism.
